The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-07-31 paper/referee/response-revision1.txt

MN-14-1011-MJ: Authors' response to the reviewer's comments
===========================================================

We are grateful to the referee for the insightful feedback, which has helped us
to improve the presentation of the data products.  In his/her report, the 
referee identified areas where additional information would (i) strengthen the 
role of the paper as a user's guide to the survey, and (ii) would lessen the 
burden on the reader to track down information from previous works.  
We appreciate the merits of the referee's suggestions in this respect, and have 
acted upon them by adding a total of 4 new pages (3 new sections, 3 new figures,
and 1 new table) to the manuscript.  Changes have been marked in bold.
We have also ~doubled the auxillary documentation that is available on our web
site as a direct consequence of the referee's suggestions, and will continue to
expand the on line documentation as user experience accumulates.

@@

In what follows we will start by explaining our response to each of the issues 
which the referee identified as high-level remarks (9 items), followed by the
referee's comments on specific sections (8 items).

High-level items
----------------

[Referee]
> 1. This manuscript presents a large amount of information about IPHAS and the preparation
> of and contents of the data products.  It also relies heavily on previous
> publications for general descriptions of the survey, particulars of the instrumentation,
> data processing algorithms and earlier data releases.   As such, it puts considerable burden 
> on the readers to track down information from many different sources.  It is
> understandable that the authors prefer to not duplicate information that is already
> available elsewhere, but it will make it much easier to understand and use the IPHAS data
> products if at least some of the general information  could be repeated in this
> manuscript.
> For example, it would be helpful to include concise summary of the IPHAS survey
> parameters and this data release in tabular form.  This summary would serve as
> a "IPHAS-At-A-Glance" for users, and coule include the survey area, dates, telescope,
> camera and survey bandpasses, etc.  Much of the information is presented in the paper,
> but widely spread over the manuscript.  Without at least a guide to finding the
> information, users will be challenged to find what they need.

[Authors' response]
Whilst revising the manuscript in response to the items that will follow below,
we have kept the referee's recommendation to lessen the burden on the reader in
mind.  In immediate response to this comment, we have extended Section 2 to 
(i) include an overview table of the survey parameters, which summarises 15 key 
properties of the survey, and (ii) included an extra figure to remind the reader
of the footprint of an IPHAS field.

@@

[Referee]
> 2. It was not immediately obvious if the images and catalogue from the Second IPHAS
> release supersede those in the 2008 Initial Data Release.  I believe the
> Second Release includes a reprocessing of all of the survey data to-date, so it
> is intended to replace the Initial Release, but that should be stated explicitly
> in the Introduction (e.g. in the 3rd paragraph).
> If my assumption is correct, there should be a summary statement about why the
> Second Release images and catalogue are in fact better products.  For example,
> better calibrations, processing algorithms, etc.

[Authors' response]
The referee correctly identified that we forgot to state that this new release 
supersedes the previous one.  We have added an explicit statement in this sense
to the third paragraph of Section 1.  We have also edited the last paragraph
of Section 3.1 to highlight the key differences between the past and the present
release.

@@

[Referee]
> 3. As stated in Section 7.2, the aim of IPHAS is to enable the discovery of
> emission line objects in the Galactic Plane.  Consequently, I was very surprised
> that no discussion of the sensitivity to line emission (e.g. H-alpha EW) was
> presented in the manuscript.  Perhaps this will be addressed further in the planned
> Barentsen et al paper, but it seems that at least some rudimentary demonstration
> of the sensitivity to line emission is a requirement for this paper.  The
> general photometric sensitivity in the three bands is discussed in section 6.2, and
> an illustration is given how to find emission-line candidates using the color-color
> diagram in Figure 18.  I strongly recommend that at least some examples be given
> of the r-H-alpha color excess as a function of line EW for a sample of objects
> with published spectra.

[Authors' response]
The referee requested a discussion on the sensitivity of the H-alpha 
photometry to line emission to be included in this work.  In response, we have
added one new section (7.4) and one new figure (Fig 21).
In this new section, which is titled "r-Halpha excess as a quantitative measure
of Halpha emission", we compare the spectroscopic H-alpha EWs for a sample of
classical Be stars from Raddi et al (2014) against the position of these
objects in the IPHAS colour-colour diagram.  A good agreement is found.
Moreover, the accompanying discussion offers a guide to the previous works on 
this topic.

@@

[Referee]
> 4. Many of the successful large surveys include a list of known caveats concerning
> released data products to help user's avoid pitfalls in their research.  I suggest that
> the authors generate a short list of the most important things about which users
> should be aware when using IPHAS data.  Users appreciate knowing how to avoid mistakes
> due to misinterpretation of known issues, and do not think less of a product because
> of them.

[Authors' response]
In response to the referee's suggestion, we have added a new Section 6.1
to alert users to the most common pitfalls.  We have also added a new
FAQ/Caveats section to our web page (www.iphas.org/dr2), which will be
updated as user experience accumulates.

[Referee]
> 5. In general, the images do not receive the same level of description in the manuscript
> as does the Catalogue.  Any guidance that can be given to users about them would be
> beneficial.  For example, at the very least include a description of the FITS image
> headers, analogous to the descriptions of the catalog columns that is given in
> the Appendix.  They should also mention that tables containing the image metadata are available at IPHAS web site.

[Authors' response]
Although the paper does intend to focus on the catalogue product, we agree with the
referee that it would be beneficial to give users additional guidance on the use 
of the images.  For this reason, we have made significant improvements to our 
web site by (i) providing a query tool to retrieve IPHAS images by coordinates, 
and (ii) provide more comprehensive documentation on the metadata, the header 
keywords, the flux calibration, and the WCS astrometric solution (see www.iphas.org/images).
Moreover, Section 6 of the paper has been edited to advertise these resources.
We note that the next release, DR3, intends to focus more strongly on our image
and image mosaic products, as the survey is approaching ~100% coverage.

@@

[Referee]
> 6. The fact that only 39% of the catalogue entries satisfy the criteria necessary to be
> classified as "reliable" begs the question "just what are the other 61% of objects."
> The authors should add more information about the general reliability of the
> catalogue with and without their criteria. For example, If a users selects
> "reliable" category sources, what should they expect the net reliability to be?
> Certainly not 100%.  How does reliability vary with source magnitude, SNR, or other 
> measurement parameters?  How much does the reliability vary with source environment
> (e.g. crowding, complex backgrounds)?
>
> Reliability can be gauged in a number of ways, such as comparisons with external,
> deeper catalogues (e.g. SDSS, UKIDSS GPS).

[Authors' response]
The referee questioned the nature of the objects which are marked as "reliable"
in the catalogue, and in particular questioned the nature of the large number of 
sources not flagged as "reliable".  Having considered this comment, we have come 
to conclude that we made a poor choice of terminology by adopting the labels 
"reliable" and "veryReliable" as quality qualifiers.
These classes, which are based on stringent quality criteria set out in 
Tables 6 and 7, aim to help users identify high-quality objects, e.g. by 
demanding a 10-sigma detection in all bands.  It goes without saying that 
there are plenty of objects which do not satisfy such strict criteria but are 
nevertheless genuine astrophysical objects.  The qualifier "reliable" might 
however confer the incorrect impression that such sources are "unreliable".
As a result, we have decided to change the terminology by renaming the class 
previously known as "reliable" into "a10" (which is short for "all-band 10-sigma"), 
and renaming "veryReliable" into "a10point" (short for "all-band 10-sigma and
perfectly point-like").  We believe this change in terminology will better serve 
as an operational description of the detection quality.

In addition to this change in terminology, we have adopted the referee's suggestion
to gauge the reliability of our objects against an external catalogue.  That is,
we have introduced a new Section 6.2, titled "Source reliability", in which the 
reliability of IPHAS sources as a function of the quality criteria is gauged 
against SDSS.  Moreover, we note that a separate paper is in an advanced state 
of preparation (Farnhill et al.), which will investigate the issues surrounding 
the completeness and reliability of our catalogue, e.g. by injecting artificial 
sources into the images and measuring their recovery rates, whilst also comparing 
source counts between SDSS and IPHAS in more detail.

@@

[Referee]
> 7. The authors have done a good job describing the photometric measurements and
> calibration accuracy.  However, there is relatively little discussion of the
> astrometric position reconstruction other than saying it is done relative to
> the 2MASS Point Source Catalogue, and there is no validation of the astrometric
> quality of the Catalogue other than to mention matching success.  Some quantification
> of the astrometric performance should be added (e.g. mean position residuals as a function
> of magnitude from comparisons with the astrometric reference stars, or external
> astrometric collections, such as UCAC4).
> Positions in the IPHAS Catalogue have been reconstructed with respect to 2MASS.
> However, the authors do not state if the proper motion of the 2MASS reference stars 
> was taken into account.  Although small, the net motion of the reference stars can
> collectively imprint on the Catalogue because the IPHAS observations were made at
> different epochs than 2MASS, in some cases as much as 12 years later.   
> Failure to do so can imprint the proper motion onto IPHAS, and also because the 
> IPHAS measurements span nearly 10 years (2003-2012), there can be secular variations.

[Authors' response]
We have edited Section 3.1 to offer the reader more details on the astrometric 
calibration.  We have also added the results of an external validation of our 
astrometry against UCAC4, which shows a mean residual of 0.09 +/- 0.07 arcsec at
the bright end of our catalogue.  Finally, we have also added a statement to specify
that no proper motion information was used in arriving at our astrometry, because
because hundreds of 2MASS positional reference stars were typically available in 
each CCD to fit the parameters of the WCS solution.

@@

[Referee]
(There does not appear to be an item number 8 in the referee report.)
> 9. The authors point out in the introduction that IPHAS will complement a number of
> new infrared surveys of the Galactic Plane.  The paper would benefit from at least
> one cross-comparison with the catalogue of one of those surveys, such as GLIMPSE
> or MIPSGAL, as an illustration of how the combination of surveys will enable
> new science.  At present, the only tie-in to the infrared surveys is the
> visual comparison between IPHAS and GLIMPSE images of one field.

[Authors' response]
In response to the referee's suggestion to include a cross-comparison between
IPHAS and an infrared survey, we have added a new section (7.3) which provides
a brief discussion on the overlap between the GLIMPSE and IPHAS catalogues.  
The discussion reminds the user that an optical survey will naturally be 
extinction-limited, by showing where sources which IPHAS and GLIMPE have in common
are positioned in a colour-colour diagram.

@@

[Referee]
> 4.a. It is stated in section 3.2 that the exposures were discarded unless they
> had 5-sigma limiting magnitudes of r>20 or i>19 and H-apha>19 mag. How were the
> limiting magnitudes for an exposure estimated?

[Authors' response]
We have added a footnote to Section 3.2 which details how the 5-sigma limiting 
magnitude is defined and estimated in our work.

[Referee]
> 4.b. Section 3.2 also specifies that an ellipticity limit of eps>0.3 was used.
> What is the justification of this limit?  For example, what is the potential impact on
> photometry and sensitivity at eps=0.25?

[Authors' response]
We have edited the third paragraph in Section 3.2 to explain that the choice
of our ellipticity limit is based on inspecting examples of high-ellipticity 
exposures, which revealed no evidence for degraded photometry up to ellipticities 
less than 0.3.  In fact we have found high ellipticities to be correlated with 
superior seeing, and therefore somewhat deeper frames.

@@

[Referee]
> 5.a. Were the atmospheric extinction coefficients of 0.09 (r) and 0.05 (i) simply
> adopted, or was there any attempt to measure them?  Is there an estimate of
> how much the extinction may have varied with time, either during a night, or
> seasonally?

[Authors' response]
The atmospheric extinction coefficients adopted are standard default values 
for La Palma.  No direct measures or corrections were attempted, in part because 
our re-calibration procedure naturally absorbs extinction variations as part of the
APASS-based zeropoint estimation.

@@

[Referee]
> 5.c. The first paragraph of section 4.2 on Global Calibration states that calibration
> errors  can be up to 0.1 magnitude in reference to the colour-colour diagram shown in
> Figure 4a. There are outlier populations in that figure that are up to 0.6 mags
> from the main locus of points.  These seem to be accurately corrected by
> the recalibration, but the text seems to understate the severity of the
> systematics.

[Authors' response]
We have edited the first paragraph of Section 4.2 to offer more detail on the
calibration errors, including the statement that we found the calibration error
to exceed 0.1 mag in 12 per cent of the fields, and 0.5 mag in 0.7 per cent.

[Referee]
> 5.d.  It is stated in section 4.2, paragraph 2 that 2MASS observed six calibration
> fields each hour.  That is not correct.  2MASS observed two calibration fields each hour.  
> However, each observation of a field consisted of six scans.

[Authors' response]
We have corrected the paragraph accordingly.  Fortunately, our mistake did not
affect the main point of the sentence, which is that 2MASS observed standard
fields more regularly than IPHAS.

@@

[Referee]
> 5.e The authors may wish to specify in Section 4.3 that the SDSS DR9 Galactic Plane data
> that were used to test the IPHAS calibration were specifically from the SEGUE program
> of SDSS II, and cite the appropriate reference for SEGUE.

[Authors' response]
We have added the appropriate citation for SEGUE (Yanny et al. 2009) to the 
first paragraph of Section 4.3.

[Referee]
> 6.a. Section 5.1
> Would it be possible to provide the aperture corrections applied to the aperture
> photometry for each image along with the image metadata?

[Authors' response]
Manpower-permitting, we intend to make recommended aperture corrections
available on our website later this year, as part of the preparations for the 
future DR3 release, which intends to focus on the image data more strongly.

@@

[Referee]
> Shouldn't the "saturated" column flag be specified per band?  Surely it must
> be possible for extremely red objects to be saturated in the i-band exposure
> but not r-band.

[Authors' response]
The saturation flag was already available for each band as part of the 
"errBits" columns, but as a result of the referee's comment, we have decided 
to make this information available in a more user-friendly way through the
introduction of three new boolean columns to the catalogue: "rSaturated", 
"iSaturated", and "haSaturated".
Likewise, we have introduced the new user-friendly columns "rDeblend",
"iDeblend", and "haDeblend" to flag blended sources on a per-band basis.
These changes have allowed the original, more obscure columns "rErrBits",
"iErrBits", and "haErrBits" to be removed from the catalogue, because the 
information they contained is now entirely captured within the six new columns.
Our catalogue files and column descriptions have been updated accordingly.

@@

[Referee]
> 6.b. Section 5.2
> Little information about the specifics of band-merging is provided other than the
> reference to Taylor 2006 and TMATCHN.  Users will be forced to track down
> the reference to understand the process.  The authors should consider giving a short
> summary of the algorithm.  Some important information that should be provided are:
> - How is confusion handled in the band merging process (e.g. what happens when
> more than one i-band detection can be matched to an r-band detection?).
> Source confusion in the band-merging process is mentioned in the final
> paragraph of section 6.3.

[Authors' response]
We have edition Section 5.2 to include a summary of the STILTS TMATCHN algorithm.
We have also included a warning that blended objects can occasionally fall victim 
to source confusion, which is a complicated problem that we have not attempted 
to resolve in this release, other than by means of clearly flagging blended objects 
in the catalogue and allowing users to refine the inter-band matching distance.
The last paragraph of Section 5.2 now reminds users of this more clearly.

@@

[Referee]
> - Is there any band-filling when a source is not detected in all bands?  If so, what
> is the algorithm used to provide the flux limits in the non-detected bands?

[Authors' response]
We do not provide flux limits in the non-detected bands at present.  This has
now been clarified in the first paragraph of Section 5.2.  We are investigating
the possibility to introduce flux limits in DR3, however.

@@

[Referee]
> 6.c. Section 5.3 - Table 4
> It is not clear what is the specific value is of the average magnitude given for each of
> the different band-detection combinations in Table 4?  Wouldn't something like the peak
> of the log N - log M curves be of more interest to provide some sense of the
> completeness and/or depth of the Catalogue?

[Authors' response]
Section 5.4 has been edited to make the role of Table 4 clearer.  The average
magnitudes have been removed from this table since they did not contribute to
the message which the Section attempts to convey.

@@

[Referee]
> 6.d. Section 5.4
> While the authors provide a thorough description of the criteria used for
> Catalogue selection, no reasons were offered for why those criteria were adopted.
> For example, were there requirements for completeness and reliability that were
> being satisfies?  If not, what were the objectives?
> In the first paragraph of section 5.4, it is stated the three criteria were
> enforced to limit the size of the catalogue.  Was this the main driver for establishing
> the top-level criteria?  It seems unusual in this day and age of massive catalogues
> that this would be a driver.

[Authors' response]
The main driver for imposing the top-level criteria is to avoid the catalogue
from being dominated by the wealth of spurious detections which the pipeline
naturally picks up at S/N < 5.  We have edited the first paragraph of Section 
5.4 to clarify this point.  For DR3, we consider following the example set by 
Spitzer/GLIMPSE to offer both a "highly reliable" as well as a  "more complete,
less reliable" version of the catalogue to the public.

@@

[Referee]
> As stated earlier in this review, some description should be offered as to the nature of
> the very large number of detections in the catalogue that are not classified as "reliable."
> It would be useful to describe them to the user, and to offer suggestions of
> how to recognize problematic detections, and also ones that may be real.

[Authors' response]
As we acknowledged during our response to "point 6", we believe that our choice
of the word "reliable" as the name of a class of high-quality objects was 
misleading. We believe to have resolved this problem by renaming the "reliable"
class into the "a10" class, which avoids conveying the incorrect impression that 
objects not in this class are spurious by default.

@@

[Referee]
> 7. Section 6.1, par. 2, sentence 1
> Should the work "accuracy" be "reliability"?

[Authors' response]
Indeed, we have corrected this mistake. Again, this is a result of our previously
incorrect usage of the term "reliable".

@@

[Referee]
> 8. Figure 11 - caption
> The reference to the left and right panels should presumably be to the top and bottom panels,
> unless the two figures will be arranged differently in the published version.

[Authors' response]
We have corrected this mistake.

---

Again, we thank the referee for the insightful feedback, which has not only
helped to benefit the DR2 paper, but will also be taken into account during the
preparation of DR3.

Kind regards,
The authors.

### 2014-07-31 paper/iphas-dr2-rev1.tex
First rev1 draft ready

\begin{abstract}
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering Galactic latitudes $|b| < 5\degr$
and longitudes $\ell=30\degr$ to 215\degr\ 
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
globally-calibrated source catalogue
derived from the survey,
providing single-epoch photometry
for 219 million unique sources
across 92 per cent of the footprint.
The observations were carried out between 2003 and 2012
at a median seeing of 1.1~arcsec (sampled at 0.33~arcsec/pixel)
and to a mean $5\sigma$-depth of 
21.2~($r$), 20.0~($i$) and 20.3~($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the global re-calibration,
and detail the construction of the new catalogue.
We show that the new calibration is accurate to
0.03~mag (rms) 
and recommend a series of quality criteria
to select accurate data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sightlines
and (ii) select \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure of the disk of the Milky Way.
This large-scale programme of observation
-- spanning a decade so far 
and using more than 300 nights in competitive open time 
at the Isaac Newton Telescope (INT) in La Palma --
aims to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century \citep[see][]{Kohoutek1999}. 
By increasing the sensitivity 
with respect to these preceding surveys 
by a factor $\sim$1000 (7~magnitudes), 
IPHAS can expand
the limited bright samples of Galactic emission line objects 
previously available into larger, deeper,
and more statistically-robust samples
that will better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007,Wesson2008,Aungwerojwit2012}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
new classical Be stars \citep{Raddi2013};
along with discoveries of new supernova remnants \citep{Sabin2013}
and new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Sabin2010, Corradi2011, Viironen2011}.

Over the years it has become apparent that the legacy of IPHAS 
will reach beyond these traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$ and $i$ broadband photometry 
alongside narrowband \ha\ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey's unique $(r-\ha,\ r-i)$ colour-colour
diagram has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband \ha\
alongside $r$ and $i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys (cf. the use of
Stromgren $uvbyH\beta$ photometry at blue wavelengths).
For such science applications to succeed, however, 
it is vital that the imaging data are transformed 
into a homogeneously-calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

@@

In \S\ref{sec:observations} we start by recapitulating the key points
of the survey observing strategy.
In \S\ref{sec:reduction} we describe the data reduction
and quality control procedures.
In \S\ref{sec:calibration} we explain the global re-calibration, in which
we draw upon the AAVSO Photometric All-Sky Survey (APASS)
and test our results against the Sloan Digital Sky Survey (SDSS).
In \S\ref{sec:catalogue} we explain how the source catalogue was compiled.
In \S\ref{sec:discussion} we discuss the properties of the catalogue
and in \S\ref{sec:demonstration} we demonstrate
the scientific exploitation of the colour/magnitude diagrams.
Finally, in \S\ref{sec:dataaccess} we discuss access
to the catalogue, an online library of reduced images and relevant source code,
The paper ends with conclusions in \S\ref{sec:conclusions} where we also outline
our future ambitions.

@@

\begin{table}
    \caption{Key properties of IPHAS DR2.}
    \label{tbl:survey}
    \begin{center}
        \begin{tabular}{ll}
        \toprule
        Property & Value \\
        \midrule
        Telescope & 2.5m Isaac Newton Telescope \\
        Instrument & Wide Field Camera \\
        Detectors & Four 2048 $\times$ 4100 pixel CCD's \\
        Pixel scale & 0.33 arcsec/pixel \\        
        Filters & $r$, $i$, \ha \\
        Filter properties & See Table~2 \\
		Magnitude system & Vega \\
        Exposure times & 30~s ($r$), 10~s ($i$), 120~s (\ha) \\
		Saturation limit & 13 ($r$), 12 ($i$), 12.5 (\ha) \\
        Detection limit (5$\sigma$, mean) & 21.2 ($r$), 20.0 ($i$), 20.3 (\ha) \\
        PSF FWHM (median) & 1.1\arcsec\ ($r$), 1.0\arcsec\ ($i$), 1.1\arcsec\ (\ha) \\
        Survey area & $\sim 1860$~deg$^2$ \\
        Footprint boundaries & $ -5\degr < b < +5\degr,\ 29\degr < \ell < 215\degr$ \\
        Observing period & August 2003 - November 2012 \\
        Website & www.iphas.org \\
        \bottomrule
        \end{tabular}
	\end{center}
\end{table}

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release,
we recap some key points {\bf below and in Table~1}.
IPHAS is an imaging survey of the Galactic Plane north of the celestial equator, 
from which photometry in Sloan $r$ and $i$ 
is extracted along with narrowband \ha. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive optical CCD photometry
of point sources in the Galactic Plane;
the footprint spans a box 
of roughly 180 by 10 degrees, 
covering Galactic latitudes $-5\degr < b < +5\degr$ 
and longitudes $30\degr < \ell < 215\degr$.

\begin{figure}
    \includegraphics[width=\linewidth]{figures/fieldpair/fieldpair.pdf} 
    \caption{
    Example footprint of a pointing and its offset position.
    Each field in the survey is accompanied by an offset field
    at 5\arcmin\ West and South
	to deal with gaps between the CCDs.
	This unit of observation is called a \emph{field pair}.
	The WFC is a mosaic of four CCDs,
	and hence a field pair is composed of eight CCD frames.
	In this example, each CCD is plotted in semi-transparent grey
	to highlight the overlap regions.
	}
    \label{fig:fieldpair}
\end{figure}

>>>>

\begin{abstract}
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering Galactic latitudes $|b| < 5\degr$
and longitudes $\ell=30\degr$ to 215\degr\ 
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
globally-calibrated source catalogue
derived from the survey,
providing single-epoch photometry
for 219 million unique sources
across 92 per cent of the footprint.
The observations were carried out between 2003 and 2012
at a median seeing of 1.1~arcsec (sampled at 0.33~arcsec/pixel)
and to a mean $5\sigma$-depth of 
21.2~($r$), 20.0~($i$) and 20.3~($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the global re-calibration,
and detail the construction of the new catalogue.
We show that the new calibration is accurate to
0.03~mag (rms) 
and recommend a series of quality criteria
to select accurate data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sightlines
and (ii) select and quantify \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure of the disk of the Milky Way.
This large-scale programme of observation
-- spanning a decade so far 
and using more than 300 nights in competitive open time 
at the Isaac Newton Telescope (INT) in La Palma --
aims to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century \citep[see][]{Kohoutek1999}. 
By increasing the sensitivity 
with respect to these preceding surveys 
by a factor $\sim$1000 (7~magnitudes), 
IPHAS can expand
the limited bright samples of Galactic emission line objects 
previously available into larger, deeper,
and more statistically-robust samples
that will better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011, Rodriguez2014}; 
new cataclysmic variables \citep{Witham2007,Wesson2008,Aungwerojwit2012}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
new classical Be stars \citep{Raddi2013};
along with discoveries of new supernova remnants \citep{Sabin2013}
and new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Sabin2010, Corradi2011, Viironen2011,Sabin2014}.

Over the years it has become apparent that the legacy of IPHAS 
will reach beyond these traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$ and $i$ broadband photometry 
alongside narrowband \ha\ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey's unique $(r-\ha,\ r-i)$ colour-colour
diagram has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013, Sale2014}. 
In effect, the availability of narrowband \ha\
alongside $r$ and $i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys (cf. the use of
Stromgren $uvbyH\beta$ photometry at blue wavelengths).
For such science applications to succeed, however, 
it is vital that the imaging data are transformed 
into a homogeneously-calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

@@

In \S\ref{sec:observations} we start by recapitulating the key points
of the survey observing strategy.
In \S\ref{sec:reduction} we describe the data reduction
and quality control procedures.
In \S\ref{sec:calibration} we explain the global re-calibration, in which
we draw upon the AAVSO Photometric All-Sky Survey (APASS)
and test our results against the Sloan Digital Sky Survey (SDSS).
In \S\ref{sec:catalogue} we explain how the source catalogue was compiled.
In \S\ref{sec:discussion} we discuss the properties of the catalogue
and in \S\ref{sec:demonstration} we demonstrate
the scientific exploitation of the colour/magnitude diagrams.
Finally, in \S\ref{sec:dataaccess} we discuss access
to the catalogue and an online library of reduced images.
The paper ends with conclusions in \S\ref{sec:conclusions} where we also outline
our future ambitions.

@@

\begin{table}
    \caption{Key properties of IPHAS DR2.}
    \label{tbl:survey}
    \begin{center}
        \begin{tabular}{ll}
        \toprule
        Property & Value \\
        \midrule
        Telescope & 2.5m Isaac Newton Telescope \\
        Instrument & Wide Field Camera (WFC) \\
        Detectors & Four 2048 $\times$ 4100 pixel CCD's \\
        Pixel scale & 0.33 arcsec/pixel \\        
        Filters & $r$, $i$, \ha \\
        Filter properties & See Table~2 \\
		Magnitude system & Vega \\
        Exposure times & 30~s ($r$), 10~s ($i$), 120~s (\ha) \\
		Saturation limit & 13 ($r$), 12 ($i$), 12.5 (\ha) \\
        Detection limit (5$\sigma$, mean) & 21.2 ($r$), 20.0 ($i$), 20.3 (\ha) \\
        PSF FWHM (median) & 1.1\arcsec\ ($r$), 1.0\arcsec\ ($i$), 1.1\arcsec\ (\ha) \\
        Survey area & $\sim 1860$~deg$^2$ \\
        Footprint boundaries & $ -5\degr < b < +5\degr,\ 29\degr < \ell < 215\degr$ \\
        Observing period & August 2003 - November 2012 \\
        Website & www.iphas.org \\
        \bottomrule
        \end{tabular}
	\end{center}
\end{table}

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release,
we recap some key points in this section {\bf (see Table~1 for a summary)}.
IPHAS is an imaging survey of the Galactic Plane north of the celestial equator, 
from which photometry in Sloan $r$ and $i$ 
is extracted along with narrowband \ha. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive optical CCD photometry
of point sources in the Galactic Plane;
the footprint spans a box 
of roughly 180 by 10 degrees, 
covering Galactic latitudes $-5\degr < b < +5\degr$ 
and longitudes $30\degr < \ell < 215\degr$.

\begin{figure}
    \includegraphics[width=\linewidth]{figures/fieldpair/fieldpair.pdf} 
    \caption{
    Example footprint of a pointing and its offset position.
    Each field in the survey is accompanied by an offset field
    at 5\arcmin\ West and South
	to deal with gaps between the CCDs.
	This unit of observation is called a \emph{field pair},
	which is observed in all three filters within a span of 10 minutes.
	The WFC is a mosaic of four CCDs,
	and hence a field pair is composed of eight CCD frames.
	In this example, each CCD is plotted as a semi-transparent grey rectangle
	to highlight the overlap regions.
	Note that the L-shaped arrangement of the CCD mosaic allows
	nearly the entire field to be captured using just two pointings,
	apart from two $\sim10\arcsec$-wide squares
	which are located where inter-CCD gaps overlap.	
	}
    \label{fig:fieldpair}
\end{figure}

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^2$ at a pixel scale of
0.33 arcsec/pixel.
To cover the Northern Plane with some overlap,
the survey area was divided into 7,635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in Dec 
and $+$5 arcmin in RA,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks.
{ \bf An example footprint of a pointing
and its offset position is shown in Fig.~\ref{fig:fieldpair}.}
Hence, the basic unit of observation
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within, typically, an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
Altogether, the survey contains 15,270 fields
grouped into 7,635 field pairs.
To achieve the desired survey depth
of 20th magnitude or fainter in each filter, 
the exposure times were set at 120~s (\ha), 
30~s ($r$) and 10~s ($i$)
in the vast majority of the survey observations.\footnote{The $r$-band exposure time was 10s instead of 30s in the first months of data taking. Since October 2010, the $i$-band exposure time 
has been increased from 10s to 20s to by-pass a sporadic exposure timing bug that affects the WFC.}
